---
title: "Excess mortality: data preparation"
author: "Radek Panczak"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
editor_options: 
  chunk_output_type: console
---


```{r setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

options(scipen=999)
set.seed(12345)

library(pacman)
p_load(knitr, rmdformats, 
       tidyverse, scales, readxl, haven, janitor, kableExtra,
       HMDHFDplus)

options(max.print = "75")
opts_chunk$set(cache = FALSE,
               prompt = FALSE,
               tidy = FALSE,
               comment = NA,
               message = FALSE,
               warning = FALSE)
opts_knit$set(width = 75)

mu <- Hmisc::markupSpecs$html
```

<!-- ----------------------------------------------------- -->

# HMD

[The Human Mortality Database](https://www.mortality.org/) data. 

```{r}
password <- readr::read_file("secrets/HMD.txt")

# pop
hmd_ch_pop_raw <- readHMDweb(CNTRY = "CHE", item ="Population" , 
                             username = "r.panczak@gmail.com", password = password,
                             fixup = FALSE)

write_rds(hmd_ch_pop_raw, "data-raw/mortality_org/hmd_ch_pop_raw.Rds")
write_dta(hmd_ch_pop_raw, "data-raw/mortality_org/hmd_ch_pop_raw.dta")

# deaths
hmd_ch_deaths_raw <- readHMDweb(CNTRY = "CHE", item = "Deaths_1x1",
                                username = "r.panczak@gmail.com", password = password,
                                fixup = FALSE)

write_rds(hmd_ch_deaths_raw, "data-raw/mortality_org/hmd_ch_deaths_raw.Rds")
write_dta(hmd_ch_deaths_raw, "data-raw/mortality_org/hmd_ch_deaths_raw.dta")
```

Available only yearly but for one year age bands & separate by sex. Time from `r min(hmd_ch_deaths_raw$Year)` to  `r max(hmd_ch_deaths_raw$Year)`.

Data is aggregated by year to match other sources. 

```{r}
hmd_ch_deaths <- hmd_ch_deaths_raw %>% 
  select(Year, Total) %>% 
  rename(year = Year) %>% 
  group_by(year) %>% 
  summarise(deaths_year = sum(Total))

write_rds(hmd_ch_deaths, "data/mortality_org/hmd_ch_deaths.Rds")
write_dta(hmd_ch_deaths, "data/mortality_org/hmd_ch_deaths.dta")
```

```{r}
ggplot(hmd_ch_deaths, aes(x = year)) +
  geom_line(aes(y = deaths_year)) +
  theme_bw() +
  xlab("") + ylab("Yearly number of deaths")
```


<!-- ----------------------------------------------------- -->

# BfS 1

Data from [Todesfälle nach Monat und Sterblichkeit seit 1803](https://www.bfs.admin.ch/bfs/de/home/statistiken/kataloge-datenbanken/daten.assetdetail.14387168.html)

```{r message=FALSE, warning=FALSE}
bfs_web_raw <- read_xlsx("data-raw/BfS/px-x-0102020206_111_20210223-102443.xlsx", 
                         skip = 2, na = "...") %>% 
  remove_empty() %>% 
  clean_names() %>% 
  rename(year = x1) %>% 
  filter(!is.na(todesfalle_total)) %>% 
  select(year:todesfalle_im_dezember)

bfs_web_deaths_yearly <- bfs_web_raw %>% 
  select(year, todesfalle_total) %>% 
  rename(deaths_year = todesfalle_total) 

p1 <- bfs_web_deaths_yearly %>% 
  uncount(12) %>% 
  group_by(year) %>% 
  mutate(month_num = row_number()) %>% 
  ungroup() 

p2 <- bfs_web_raw %>% 
  select(-todesfalle_total) %>% 
  filter(!is.na(todesfalle_im_januar)) %>%
  pivot_longer(!year, names_to = "month", values_to = "deaths_month") %>% 
  mutate(month = str_replace(month, fixed("todesfalle_im_"), "")) %>% 
  group_by(year) %>% 
  mutate(month_num = row_number()) %>% 
  ungroup() %>% 
  select(-month) 

bfs_web_deaths_monthly <- p1 %>% 
  left_join(p2) %>% 
  mutate(date = anytime::anydate(paste0(year, "-", month_num))) %>% 
  relocate(year, month_num, date) %>% 
  mutate(type = if_else(is.na(deaths_month), "yearly", "monthly"))

write_rds(bfs_web_deaths_monthly, "data/BfS/bfs_web_deaths_monthly.Rds")
write_dta(bfs_web_deaths_monthly, "data/BfS/bfs_web_deaths_monthly.dta")
```

Yearly deaths for period `r min(bfs_web_deaths_yearly$year)` to  `r max(bfs_web_deaths_yearly$year)`; monthly deaths for period 1901 to `r max(p2$year)`, with special addition of period `r min(p2$year)` to 1885.

```{r}
ggplot(data = filter(bfs_web_deaths_monthly, 
                     type == "yearly" &
                       year < 1877), 
       aes(x = date)) +
  geom_line(aes(y = deaths_year)) +
  geom_line(data = filter(bfs_web_deaths_monthly, 
                          type == "yearly" &
                            year > 1885), 
            aes(y = deaths_year)) +
  geom_line(data = filter(bfs_web_deaths_monthly, 
                          type == "monthly" & 
                            year >= 1877 & year <= 1885), 
            aes(y = deaths_month)) +
  geom_line(data = filter(bfs_web_deaths_monthly, 
                          type == "monthly" & 
                            year >= 1901), 
            aes(y = deaths_month)) +
  facet_grid(vars(type), scales = "free_y") +
  theme_bw() +
  xlab("") + ylab("Monthly/yearly number of deaths")
```

Note different `y` scales! 

```{r}
inconsistent <- bfs_web_deaths_monthly %>% 
  filter(!is.na(deaths_month)) %>% 
  group_by(year) %>% 
  mutate(deaths_summed = sum(deaths_month)) %>% 
  ungroup() %>% 
  filter(deaths_year != deaths_summed)
```

There are three years where summed monthly data differ from 'official' yearly sums.

```{r}
inconsistent %>% 
  filter(month_num == 1) %>% 
  select(year, deaths_year, deaths_summed) %>% 
  mutate(difference = deaths_year - deaths_summed) %>% 
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>% 
  column_spec(1, italic = TRUE) %>%
  row_spec(1, italic = FALSE)
```

<!-- ----------------------------------------------------- -->

# Computing Environment

`r mu$session()`