---
title: "Excess mortality"
subtitle: "Data preparation"
author: "Radek Panczak"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
---

<!-- ----------------------------------------------------- -->

```{r setup, include = FALSE}
set.seed(12345)
options(scipen = 999)
options(max.print = "75")

library(pacman)
p_load(tidyverse, magrittr, janitor, scales,
       readxl, lubridate, aweek, wktmo,
       kableExtra, lemon, skimr)

import::from("pxR", "read.px")
# import::from("lemon", "scale_x_symmetric")
# import::from("wktmo", "weekToMonth")
```

```{r knit-setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_chunk$set(cache = FALSE,
                      prompt = FALSE,
                      tidy = FALSE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE)

knitr::opts_knit$set(width = 75)

mu <- Hmisc::markupSpecs$html
```

<!-- ----------------------------------------------------- -->

# COVID-19 data

Pulling data directly from [Johns Hopkins Coronavirus Resource Center](https://coronavirus.jhu.edu/about/how-to-use-our-data) [github repo](https://github.com/CSSEGISandData/COVID-19) (using [`csse_covid_19_daily_reports`](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_daily_reports) with appropriate time stamps).   

Sources of data given for three analysed countries are:  

- Spain - [RTVE](https://www.rtve.es/noticias/20200514/mapa-del-coronavirus-espana/2004681.shtml)
- Sweden - [The Swedish Public Health Agency](https://experience.arcgis.com/experience/09f821667ce64bf7be6f9f87457ed9aa)
- Switzerland 
- [Federal Office Of Public Health](https://www.bag.admin.ch/bag/en/home/krankheiten/ausbrueche-epidemien-pandemien/aktuelle-ausbrueche-epidemien/novel-cov/situation-schweiz-und-international.html)
- [Open Government Data Reported By The Swiss Cantons](https://github.com/openZH/covid_19)

```{r}
covid_cases_deaths_jh_20 <-   read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/12-31-2020.csv") %>% 
  filter(Country_Region %in% c("Switzerland", "Sweden", "Spain")) %>% 
  group_by(Country_Region) %>% 
  summarise(Confirmed = sum(Confirmed, na.rm = TRUE),
            Deaths = sum(Deaths, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(Year = 2020)

covid_cases_deaths_jh <- bind_rows(
  
  covid_cases_deaths_jh_20,
  
  read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/07-01-2021.csv") %>% 
    filter(Country_Region %in% c("Switzerland", "Sweden", "Spain")) %>% 
    group_by(Country_Region) %>% 
    summarise(Confirmed = sum(Confirmed, na.rm = TRUE),
              Deaths = sum(Deaths, na.rm = TRUE)) %>% 
    ungroup() %>% 
    left_join(covid_cases_deaths_jh_20 %>% 
                rename (Deaths_20 = Deaths,
                        Confirmed_20 = Confirmed)) %>% 
    mutate(Deaths = Deaths - Deaths_20,
           Confirmed = Confirmed - Confirmed_20) %>% 
    select(-ends_with("_20")) %>% 
    mutate(Year = 2021)
) %>% 
  rename(Country = Country_Region) %>% 
  select(-Confirmed)

write_rds(covid_cases_deaths_jh, "data/covid_cases_deaths_jh.Rds")
```

```{r include=FALSE}
rm(covid_cases_deaths_jh_20)
```

<!-- ----------------------------------------------------- -->

# HMD - yearly data with age & sex structure

[The Human Mortality Database](https://www.mortality.org/) is the source of the data.  

Available on yearly resolution only but for one year age bands & separate by sex. 

The dataset has been last updated on `2021-11-10`.  

In order to update newer data the [chunk option](https://yihui.org/knitr/options/) of the code snippet below has to be changed to `eval=TRUE` and a password to user account on mortality.org website provided (stored in plain text file `HMD.txt` in `secrets` directory of the project).  

```{r eval=FALSE}
p_load(HMDHFDplus)

# text file with password or modify password in calls below
password <- readr::read_file("secrets/HMD.txt")

# CH
# pop
hmd_ch_pop_raw <- readHMDweb(CNTRY = "CHE", item ="Population" , 
                             username = "r.panczak@gmail.com", password = password,
                             fixup = FALSE) %>% 
  as_tibble()

write_rds(hmd_ch_pop_raw, "data-raw/mortality_org/hmd_ch_pop_raw.Rds")

# deaths
hmd_ch_deaths_raw <- readHMDweb(CNTRY = "CHE", item = "Deaths_1x1",
                                username = "r.panczak@gmail.com", password = password,
                                fixup = FALSE) %>% 
  as_tibble()

write_rds(hmd_ch_deaths_raw, "data-raw/mortality_org/hmd_ch_deaths_raw.Rds")

# ES
# pop
# 1975 problem resolved in issue #45
hmd_es_pop_raw <- readHMDweb(CNTRY = "ESP", item ="Population" , 
                             username = "r.panczak@gmail.com", password = password,
                             fixup = FALSE) %>% 
  as_tibble() %>% 
  mutate(Year = ifelse(Year == "1975+", 1975, Year)) %>% 
  filter(Year != "1975-") %>% 
  mutate(Year = as.numeric(Year))

write_rds(hmd_es_pop_raw, "data-raw/mortality_org/hmd_es_pop_raw.Rds")

# deaths
hmd_es_deaths_raw <- readHMDweb(CNTRY = "ESP", item = "Deaths_1x1",
                                username = "r.panczak@gmail.com", password = password,
                                fixup = FALSE) %>% 
  as_tibble()

write_rds(hmd_es_deaths_raw, "data-raw/mortality_org/hmd_es_deaths_raw.Rds")

# SE
# pop
hmd_se_pop_raw <- readHMDweb(CNTRY = "SWE", item ="Population" , 
                             username = "r.panczak@gmail.com", password = password,
                             fixup = FALSE) %>% 
  as_tibble()

write_rds(hmd_se_pop_raw, "data-raw/mortality_org/hmd_se_pop_raw.Rds")

# deaths
hmd_se_deaths_raw <- readHMDweb(CNTRY = "SWE", item = "Deaths_1x1",
                                username = "r.panczak@gmail.com", password = password,
                                fixup = FALSE) %>% 
  as_tibble()

write_rds(hmd_se_deaths_raw, "data-raw/mortality_org/hmd_se_deaths_raw.Rds")

p_unload(HMDHFDplus)
```

## Yearly deaths

Combining three countries into one dataset.  

```{r}
hmd_deaths_age_sex <- 
  bind_rows(
    read_rds("data-raw/mortality_org/hmd_es_deaths_raw.Rds") %>% mutate(Country = "Spain"), 
    read_rds("data-raw/mortality_org/hmd_se_deaths_raw.Rds") %>% mutate(Country = "Sweden"), 
    read_rds("data-raw/mortality_org/hmd_ch_deaths_raw.Rds") %>% mutate(Country = "Switzerland")
  ) %>% 
  select(-Male, -Female) %>% 
  mutate(Age = if_else(Age == "110+", "110", Age)) %>% 
  mutate(Year = as.integer(Year),
         Age = as.integer(Age),
         Deaths = as.integer(round(Total))) %>% 
  select(-Total) %>% 
  relocate(Country) 

write_rds(hmd_deaths_age_sex, "data/mortality_org/hmd_deaths_age_sex.Rds")
```

Note: Spain has *two years less* of data than other two countries:  

```{r echo=FALSE}
hmd_deaths_age_sex %>% 
  group_by(Country) %>% 
  summarize(Min = min(Year),
            Max = max(Year)) 
```

Yearly totals per country indicating different time frames of data.  

```{r}
hmd_deaths_year <- hmd_deaths_age_sex %>% 
  group_by(Country, Year) %>%
  summarise(Deaths = as.integer(round(sum(Deaths)))) %>% 
  ungroup()

write_rds(hmd_deaths_year, "data/mortality_org/hmd_deaths_year.Rds")
```

```{r echo=FALSE}
hmd_deaths_year %>%
  ggplot(aes(x = Year)) +
  geom_line(aes(y = Deaths, color = Country)) +
  theme_bw() +
  xlab("") + ylab("Yearly number of deaths")
```

Example of age breakdown, showing demographic transition and first, crude indication of the effects of pandemics (and also - smaller coverage of ESP data!).  

```{r echo=FALSE}
hmd_deaths_age_sex %>% 
  filter(Year %in% c(1876, 1917, 1918, 2019, 2020)) %>% 
  mutate(Age_cat = cut(Age, c(seq(0, 90, 10), 120), 
                       include.lowest = TRUE, right = FALSE)) %>% 
  group_by(Country, Year, Age_cat) %>% 
  summarise(Deaths = sum(Deaths)) %>% 
  ungroup() %>% 
  ggplot(mapping = aes(x = Deaths, 
                       y = Age_cat)) +
  geom_col() +
  labs(x = "Deaths", y = "Age at death") +
  facet_grid(vars(Year), vars(Country), scales = "free_x") +
  theme_bw() 
```

## Yearly population

Similarly, population data is available with 1y age bands by year.  
```{r include=FALSE}
hmd_pop_age_sex <- 
  bind_rows(
    read_rds("data-raw/mortality_org/hmd_es_pop_raw.Rds") %>%
      mutate(Country = "Spain"),
    read_rds("data-raw/mortality_org/hmd_se_pop_raw.Rds") %>%
      mutate(Country = "Sweden"),
    read_rds("data-raw/mortality_org/hmd_ch_pop_raw.Rds") %>%
      mutate(Country = "Switzerland")
  ) %>% 
  select(-Male, -Female) %>% 
  mutate(Age = if_else(Age == "110+", "110", Age)) %>% 
  mutate(Age = as.integer(Age),
         Year = as.integer(Year),
         Population = as.integer(round(Total))) %>% 
  select(-Total) %>% 
  relocate(Country) 

write_rds(hmd_pop_age_sex, "data/mortality_org/hmd_pop_age_sex.Rds")
```

Again, *Spain has less data*:  

```{r echo=FALSE}
hmd_pop_age_sex %>% 
  group_by(Country) %>% 
  summarize(Min = min(Year),
            Max = max(Year)) 
```

Yearly totals  

```{r}
hmd_pop_year <- hmd_pop_age_sex %>% 
  group_by(Country, Year) %>%
  summarise(Population = as.integer(round(sum(Population)))) %>% 
  ungroup()

write_rds(hmd_pop_year, "data/mortality_org/hmd_pop_year.Rds")
```

```{r echo=FALSE}
hmd_pop_year  %>% 
  ggplot(aes(x = Year)) +
  geom_line(aes(y = Population/1000000, colour = Country)) +
  theme_bw() +
  xlab("") + ylab("Yearly population [million]") +
  scale_y_continuous(labels = comma)
```

Again, age structure is available on yearly resolution. Example of three countries (starting in `1908` - at minimum year of coverage for all countries).      

```{r echo=FALSE}
hmd_pop_age_sex %>% 
  filter(Year %in% c(1908, 2019)) %>% 
  mutate(Age_cat = cut(Age, c(seq(0, 90, 10), 120), 
                       include.lowest = TRUE, right = FALSE)) %>% 
  group_by(Country, Year, Age_cat) %>% 
  summarise(Population = sum(Population)) %>% 
  ungroup() %>% 
  ggplot(mapping = aes(x = Population/1000000, 
                       y = Age_cat)) +
  geom_col() +
  labs(x = "Population [million]", y = "Age") +
  facet_grid(vars(Year), vars(Country), scales = "free_x") +
  theme_bw()
```

<!-- ----------------------------------------------------- -->

# Switzerland 

## Monthly deaths until 2020

Historical data from [Todesfälle nach Monat und Sterblichkeit seit 1803](https://www.bfs.admin.ch/bfs/de/home/statistiken/kataloge-datenbanken/daten.assetdetail.18584915.html).  

Using data **only since** `1877` when *monthly reporting starts*.  

```{r message=FALSE, warning=FALSE}
ch_deaths_month <- read.px("data-raw/BfS/px-x-0102020206_111.px") %>% 
  as_tibble() %>% remove_empty() %>% clean_names() %>% 
  rename(Year = jahr,
         Deaths = value) %>% 
  mutate(Year = as.numeric(as.character(Year))) %>% 
  filter(Year >= 1877) %>% 
  filter(str_detect(demografisches_merkmal_und_indikator, 'Todesfälle im')) %>% 
  droplevels(.) %>% 
  mutate(Month = rep(seq(1:12), times = nrow(.)/12) ) %>% 
  select(-demografisches_merkmal_und_indikator) %>% 
  mutate(Date = ymd(paste0(Year, "-", Month, "-01")),
         Year = as.integer(Year),
         Deaths = as.integer(Deaths)) %>% 
  relocate(Year, Month, Date)
```

## Monthly 2021 deaths 

BfS delivered custom data to supplement first six months of `2021`. Merged to dataset covering period until `2020`.  

```{r}
ch_deaths_2021 <- read_csv("data-raw/BfS/Lieferung_Staub_2021_MonatsdatenJANbisJUN_Update04NOV21.csv", col_types = cols(TodJahr = col_integer(), TodMonat = col_integer(), Alter = col_integer(), COUNT = col_integer()), trim_ws = TRUE) %>% 
  rename(Year = TodJahr, Month = TodMonat, Deaths = COUNT) %>% 
  filter(Year == as.integer(2021)) %>% 
  group_by(Year, Month) %>% 
  summarise(Deaths = sum(Deaths)) %>% 
  ungroup() %>% 
  mutate(Date = ymd(paste0(Year, "-", Month, "-01")),
         Deaths = as.integer(round(Deaths))) %>% 
  select(Year, Month, Date, Deaths)
```

```{r include=FALSE}
ch_deaths_month %<>% 
  bind_rows(ch_deaths_2021) %>% 
  mutate(Country = "Switzerland") %>% 
  relocate(Country)

write_rds(ch_deaths_month, "data/BfS/ch_deaths_month.Rds")

rm(ch_deaths_2021)
```

## Monthly datasets combined 

```{r echo=FALSE}
ggplot(data = ch_deaths_month, 
       aes(x = Date)) +
  geom_line(aes(y = Deaths / 1000)) +
  theme_bw() +
  xlab("") + ylab("Monthly number of deaths [thousand]")
```

## Yearly 2021 deaths with age & sex structure

BfS delivered data to supplement missing six months of `2021`.  

```{r}
ch_deaths_age_sex_2021 <- read_csv("data-raw/BfS/Lieferung_Staub_2021_MonatsdatenJANbisJUN_Update04NOV21.csv", col_types = cols(TodJahr = col_integer(), TodMonat = col_integer(), Alter = col_integer(), COUNT = col_integer()), trim_ws = TRUE) %>% 
  rename(Year = TodJahr, Age = Alter, Sex = Geschlecht, Deaths = COUNT) %>% 
  select(-TodMonat) %>% 
  group_by(Year, Age) %>% 
  summarise(Deaths = sum(Deaths)) %>% 
  ungroup() %>% 
  mutate(Country = "Switzerland",
         Deaths = as.integer(round(Deaths))) %>% 
  relocate(Country)
```

<!-- ----------------------------------------------------- -->

# Spain

## Monthly deaths untill 2019

Data from INE:
- data prior to 1940 was transcribed from PDFs available from INE  
- *Vital Statistics: Deaths (Historical series 1941-1974) `1941-1974` [here](https://www.ine.es/jaxiT3/Tabla.htm?t=6563&L=1)  
- *Vital Statistics: Deaths (Series since 1975)* since `1975` [here](https://www.ine.es/jaxiT3/Tabla.htm?t=6561&L=1)  

```{r}
es_deaths_month <- read_xlsx("data-raw/INE/Spain.xlsx") %>% 
  remove_empty(which = c("rows", "cols")) %>% 
  clean_names() %>% 
  rename(Year = jahr) %>% 
  pivot_longer(!Year, names_to = "Month", values_to = "Deaths") %>% 
  mutate(Month = as.integer(str_replace(Month, fixed("m"), "")),
         Year = as.integer(Year),
         Deaths = as.integer(round(Deaths)))  %>%
  mutate(Date = ymd(paste0(Year, "-", Month, "-01"))) %>% 
  relocate(Year, Month, Date) %>% 
  filter(Year != 2020)
```

## Monthly 2020 deaths 

`2020` update is from *Vital Statistics. **Provisional** data: Year 2020* [here](https://www.ine.es/jaxi/Tabla.htm?path=/t20/e301/provi/l1/&file=02001.px&L=1)

```{r}
source("R/fn_get_death_data_spain.R")

es_2020 <- fn_get_death_data_spain("data-raw/INE/Spain_2020.xlsx") %>%
  gather(., Month, Deaths, January:December, factor_key = TRUE) %>%
  group_by(Month) %>%
  summarize(Deaths = sum(Deaths, na.rm = TRUE)) %>%
  ungroup() %>% 
  mutate(Year = as.integer(2020),
         Deaths = as.integer(Deaths),
         Date = ymd(paste0(Year, Month, "-01")),
         Month = as.integer(Month)) %>% 
  relocate(Year, Month, Date, Deaths)
```

## Monthly 2021 deaths 

`2021` data is from *Estimate of number of weekly deaths during the Covid-19 outbreak* [here](https://www.ine.es/en/experimental/defunciones/experimental_defunciones.htm#tablas_resultados).  

Note: data are in *weekly format* and **converted** with methods (and caveats!) described STMF chapter of weekly data doc.  

```{r}
temp <- read_xlsx("data-raw/INE/Spain_2021_death_up.xlsx", skip = 9) %>%   
  remove_empty(which = c("rows", "cols")) %>% 
  clean_names() %>% 
  rename(Week = x1) %>% 
  filter(!is.na(from_0_to_4_years_old)) %>% 
  mutate(Deaths = rowSums(.[-1])) %>% 
  select(Week, Deaths)

es_2021 <- weekToMonth(rev(temp$Deaths), datStart = "31-12-2018") %>% 
  as_tibble() %>% 
  separate(yearMonth, c("Year", "Month"), convert = TRUE) %>% 
  rename(Deaths = value) %>% 
  mutate(Date = ymd(paste0(Year, "-", Month, "-01")),
         Deaths = as.integer(Deaths)) %>% 
  relocate(Year, Month, Date) %>% 
  filter(Year == 2021) %>% 
  filter(Month <= 6)
```

## Monthly datasets combined 

```{r include=FALSE}
es_deaths_month %<>% 
  bind_rows(es_2020) %>% 
  bind_rows(es_2021) %>% 
  mutate(Country = "Spain") %>% 
  relocate(Country)

rm(es_2020, es_2021, temp); gc()
```

```{r echo=FALSE}
es_deaths_month %>% 
  ggplot(aes(x = Date, y = Deaths)) +
  geom_line() +
  theme_bw() +
  xlab("") + ylab("Monthly deaths")
```

## Yearly 2019-2020 deaths with age structure

INE data to supplement missing 2019-20 years in HMD.  

`2019` data available in *Vital Statistics: Deaths* [here](https://www.ine.es/jaxiT3/Tabla.htm?t=6547&L=1) and `2020` data available in*Vital Statistics. Provisional data: Year 2020*  [here](https://www.ine.es/jaxi/Tabla.htm?path=/t20/e301/provi/l1/&file=02001.px&L=1).  

```{r}
es_deaths_age_2019_2020 <- bind_rows(
  
  read_delim("data-raw/INE/Spain2019age.csv",
             delim = ";",
             locale = locale(grouping_mark = "."),
             trim_ws = TRUE) %>% 
    mutate(Year = as.integer(2019)) %>% 
    mutate(Age = word(Edad, 1),
           Age = str_replace(Age, fixed("Menores"), "0"),
           Age = as.integer(Age)
    ) %>% 
    select(Year, Age, Total) %>% 
    mutate(Deaths = as.integer(round(Total))) %>% 
    select(-Total), 
  
  fn_get_death_data_spain("data-raw/INE/Spain_2020.xlsx") %>%
    gather(., Month, Deaths, January:December, factor_key = TRUE) %>%
    group_by(Age) %>%
    dplyr::summarize(Deaths = sum(Deaths, na.rm = TRUE)) %>%
    mutate(Year = as.integer(2020),
           Deaths = as.integer(Deaths)) %>%
    select(Year, Age, Deaths)
  
) %>% 
  mutate(Country = "Spain") %>% relocate(Country)
```

## Yearly 2021 deaths with age structure

INE data to supplement missing 6 months of `2021` in HMD available from *Weekly death estimates (EDeS) during the covid-19 outbreak* [here](https://www.ine.es/en/experimental/defunciones/experimental_defunciones.htm#tablas_resultados).  

Note: data are in weekly format and converted with methods (and caveats!) described above in STMF chapter.  

```{r}
temp <- read_xlsx("data-raw/INE/Spain_2021_death_up.xlsx", skip = 9) %>%   
  remove_empty(which = c("rows", "cols")) %>% 
  clean_names() %>% 
  rename(Week = x1) %>% 
  filter(!is.na(from_0_to_4_years_old))

es_deaths_age_2021 <- bind_rows(
  
  weekToMonth(rev(temp$from_0_to_4_years_old), datStart = "31-12-2018") %>% 
    mutate(Age = 0) %>% as_tibble(),
  
  weekToMonth(rev(temp$from_5_to_9_years), datStart = "31-12-2018") %>% 
    mutate(Age = 5) %>% as_tibble(),
  
  weekToMonth(rev(temp$from_10_to_14_years), datStart = "31-12-2018") %>% 
    mutate(Age = 10) %>% as_tibble(),
  
  weekToMonth(rev(temp$from_15_to_19_years), datStart = "31-12-2018") %>% 
    mutate(Age = 15) %>% as_tibble(),
  
  weekToMonth(rev(temp$from_20_to_24_years), datStart = "31-12-2018") %>% 
    mutate(Age = 20) %>% as_tibble(),
  
  weekToMonth(rev(temp$from_25_to_29_years), datStart = "31-12-2018") %>% 
    mutate(Age = 25) %>% as_tibble(),
  
  weekToMonth(rev(temp$from_30_to_34_years_old), datStart = "31-12-2018") %>% 
    mutate(Age = 30) %>% as_tibble(),
  
  weekToMonth(rev(temp$from_35_to_39_years_old), datStart = "31-12-2018") %>% 
    mutate(Age = 35) %>% as_tibble(),
  
  weekToMonth(rev(temp$from_40_to_44_years), datStart = "31-12-2018") %>% 
    mutate(Age = 40) %>% as_tibble(),
  
  weekToMonth(rev(temp$from_45_to_49_years_old), datStart = "31-12-2018") %>% 
    mutate(Age = 45) %>% as_tibble(),
  
  weekToMonth(rev(temp$from_50_to_54_years_old), datStart = "31-12-2018") %>% 
    mutate(Age = 50) %>% as_tibble(),
  
  weekToMonth(rev(temp$from_55_to_59_years_old), datStart = "31-12-2018") %>% 
    mutate(Age = 55) %>% as_tibble(),
  
  weekToMonth(rev(temp$from_60_to_64_years_old), datStart = "31-12-2018") %>% 
    mutate(Age = 60) %>% as_tibble(),
  
  weekToMonth(rev(temp$from_65_to_69_years_old), datStart = "31-12-2018") %>% 
    mutate(Age = 65) %>% as_tibble(),
  
  weekToMonth(rev(temp$from_70_to_74_years), datStart = "31-12-2018") %>% 
    mutate(Age = 70) %>% as_tibble(),
  
  weekToMonth(rev(temp$from_75_to_79_years), datStart = "31-12-2018") %>% 
    mutate(Age = 75) %>% as_tibble(),
  
  weekToMonth(rev(temp$from_80_to_84_years), datStart = "31-12-2018") %>% 
    mutate(Age = 80) %>% as_tibble(),
  
  weekToMonth(rev(temp$from_85_to_89_years), datStart = "31-12-2018") %>% 
    mutate(Age = 85) %>% as_tibble(),
  
  weekToMonth(rev(temp$x90_years_old_and_over), datStart = "31-12-2018") %>% 
    mutate(Age = 90) %>% as_tibble()
  
) %>% 
  as_tibble() %>% 
  mutate(Country = "Spain") %>%
  separate(yearMonth, c("Year", "Month"), convert = TRUE) %>% 
  rename(Deaths = value) %>% 
  mutate(Deaths = as.integer(Deaths),  
         Age = as.integer(Age)) %>% 
  relocate(Country, Year, Month, Age) %>% 
  filter(Year == 2021) %>% 
  filter(Month <= 6) %>% 
  group_by(Country, Year, Age) %>% 
  summarise(Deaths = sum(Deaths)) %>% 
  ungroup()

rm(temp)
```

<!-- ----------------------------------------------------- -->

# Sweden

## Monthly deaths untill 2020 

```{r}
se_deaths_month <- read_csv("data-raw/SCB/000000NF_20210315-164930.csv",
                            skip = 2) %>% 
  remove_empty(which = c("rows", "cols")) %>% 
  clean_names() %>% 
  rename(Month_txt = month) %>% 
  filter(Month_txt != "Unknown") %>% 
  select(-sex) %>% 
  pivot_longer(!Month_txt, names_to = "Year", values_to = "Deaths") %>% 
  group_by(Year, Month_txt) %>% 
  summarise(Deaths = sum(Deaths)) %>% 
  ungroup() %>% 
  mutate(Year = as.integer(str_replace(Year, fixed("x"), "")),
         Month = match(Month_txt, month.name),
         Deaths = as.integer(round(Deaths))) %>% 
  select(-Month_txt) %>% 
  mutate(Date = ymd(paste0(Year, "-", Month, "-01"))) %>% 
  relocate(Year, Month, Date) %>% 
  arrange(Year, Month)
```

## Monthly 2021 deaths

[Preliminär statistik över döda (Excel)](https://www.scb.se/hitta-statistik/sverige-i-siffror/manniskorna-i-sverige/doda-i-sverige/) file is available from SCB.  Data are available till October.   

```{r eval=FALSE}
# curl command doesn't work with this URL
# please download manually using link above!
download.file(url = "https://www.scb.se/hitta-statistik/statistik-efter-amne/befolkning/befolkningens-sammansattning/befolkningsstatistik/pong/tabell-och-diagram/preliminar-statistik-over-doda/",
              destfile = "data-raw/SCB/2021-08-23-preliminar_statistik_over_doda_inkl_eng.xlsx",
              method = "curl")
```

```{r}
se_deaths_month_2021 <- read_excel("data-raw/SCB/2021-11-01-preliminar_statistik_over_doda_inkl_eng.xlsx", sheet = "Tabell 10", skip = 10) %>% 
  filter(Region == "Hela riket" & Period == "2021 antal") %>% 
  select(Januari:Oktober) %>% 
  pivot_longer(Januari:Oktober) %>% 
  mutate(Year = as.integer(2021),
         Month = 1:10,
         Date = ymd(paste0(Year, "-", Month, "-01")),
         Deaths = as.integer(value)
  ) %>% 
  select(-name, -value) %>% 
  filter(Month <= 6)

se_deaths_month %<>% 
  bind_rows(se_deaths_month_2021) %>% 
  mutate(Country = "Sweden") %>% relocate(Country)
```

```{r include=FALSE}
rm(se_deaths_month_2021)
```

## Monthly deaths combined 

```{r echo=FALSE}
se_deaths_month %>% 
  ggplot(aes(x = Date, y = Deaths)) +
  geom_line() +
  theme_bw() +
  xlab("") + ylab("Monthly deaths")
```

<!-- ----------------------------------------------------- -->

# Projected 2020-21 population numbers {.tabset}

Data used for projection come from HMD.  

```{r}
hmd_pop_age_agg <- hmd_pop_age_sex %>% 
  mutate(Age_cat = cut(Age, c(seq(0, 90, 10), 120),
                       include.lowest = TRUE, right = FALSE)) %>% 
  group_by(Country, Year, Age_cat) %>% 
  summarise(Population = sum(Population)) %>% 
  ungroup()
```

```{r include=FALSE}
rm(hmd_pop_age_sex)
```

As explained above, Spain has less data:  

```{r}
hmd_pop_age_agg %>% 
  # filter(Year >= 2010 & Year <= 2019) %>%
  filter(Year >= 2015) %>%
  ggplot(
    aes(x = Year, y = Population, fill = factor(Year))) +
  geom_col() +
  scale_y_continuous() +
  theme_bw() +
  scale_fill_viridis_d(option = "E", direction = -1) +
  xlab("Age group") + ylab("Population difference between sources") + 
  facet_grid(vars(Country), vars(Age_cat), scales = "free_y") +
  geom_vline(xintercept = 2019.5, color = "red")
```

For each year, age, country stratum, linear model is fit using the `2015-2019` data; then prediction is made for `2020` & `2021` years. The gist of this operation is to exclude the effect of pandemic from the population counts in years affected.   

```{r}
# p_load(mgcv)
# p_load(splines)
# knots <- quantile(seq.int(2010, 2019), p = c(0.25, 0.5, 0.75))

pop_age_2020 <- hmd_pop_age_agg %>% 
  filter(Year >= 2015 & Year <= 2019) %>% 
  group_by(Country, Age_cat) %>% 
  # do(broom::tidy(lm(population ~ year, data = .))) %>% 
  # do(broom::augment(lm(population ~ year, data = .))) %>% 
  # lm
  do(lm(Population ~ Year, data = .) %>%
       # polynomial
       # do(lm(Population ~ poly(Year, 2, raw = TRUE), data = .) %>%
       # spline
       # do(lm(Population ~ bs(Year, knots = knots), data = .) %>%
       # GAM     
       # do(gam(Population ~ s(Year), data = .) %>%
       predict(., newdata = tibble(Year = c(2020, 2021))) %>%
       tibble(Year = c(2020, 2021), 
              Population = .)
  ) %>% 
  ungroup() %>% 
  mutate(Population = as.integer(round(Population)),
         Year = as.integer(round(Year))) %>% 
  arrange(Country, Year, Age_cat) %>% 
  relocate(Country, Year, Age_cat, Population)
```

For comparison against `2020` & `2021` we use data from HMD (for SWE & CHE) and INE (for ESP). For the latter we use `Main series since 1971` from [here](https://www.ine.es/jaxiT3/Tabla.htm?t=31304). # HMD docs say that pop Spain 2013-2019 are from: *"1 Jan"* so using data from this time point for consistency.  

```{r}
es_pop_age_new <- 
  bind_rows(
    read_excel("data-raw/INE/Spain_Population_age-2019.xlsx",
               sheet=1,range="A8:H212") %>%
      slice(., -(1:2)) %>%
      select(1,8) %>%
      rename(Population=`Both sexes...8`) %>%
      filter(!is.na(Population)) %>%
      mutate(Age=0:(nrow(.)-1)) %>%
      select(Age, Population) %>% 
      mutate(Year = as.integer(2020)), 
    
    read_excel("data-raw/INE/Spain_Population_age-2019.xlsx",
               sheet=1,range="A8:B212") %>%
      slice(., -(1:2)) %>%
      rename(Population=`Both sexes`) %>%
      filter(!is.na(Population)) %>%
      mutate(Age=0:(nrow(.)-1)) %>%
      select(Age, Population) %>% 
      mutate(Year = as.integer(2021))
  ) %>% 
  mutate(Age_cat = cut(Age, c(seq(0, 90, 10), 120),
                       include.lowest = TRUE, right = FALSE
  )) %>% 
  group_by(Year, Age_cat) %>% 
  summarise(Population = sum(Population)) %>% 
  ungroup() %>% 
  mutate(Population = as.integer(Population),
         Country = "Spain") %>% 
  relocate(Country, Year)
```

Absolute difference across countries and age groups in `2020/21`:  

```{r echo=FALSE, warning=FALSE}
bind_rows(
  
  hmd_pop_age_agg %>% 
    filter(Year >= 2020), 
  
  es_pop_age_new
  
) %>% 
  mutate(Source = "HMD") %>% 
  bind_rows(pop_age_2020 %>% 
              mutate(Source = "Estimate")) %>% 
  ggplot(
    aes(x = Age_cat, y = Population / 1000000, 
        fill = factor(Source))) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +
  theme_bw() +
  scale_fill_viridis_d(option = "E", direction = 1) +
  xlab("Age group") + ylab("Yearly population [millions]") + 
  facet_grid(vars(Country), vars(Year), scales = "free_y")
```

Relative differences in `2020/21`:  

```{r echo=FALSE, warning=FALSE}
bind_rows(
  
  hmd_pop_age_agg %>% 
    filter(Year >= 2020), 
  
  es_pop_age_new
  
) %>% 
  rename(Population_hmd = Population) %>% 
  full_join(pop_age_2020 %>% 
              rename(Population_lm = Population)) %>% 
  mutate(Difference = (Population_lm - Population_hmd) / Population_lm) %>% 
  ggplot(
    aes(x = Age_cat, y = Difference)) +
  geom_col() +
  scale_y_continuous(labels = percent) +
  theme_bw() +
  scale_fill_viridis_d(option = "E", direction = 1) +
  xlab("Age group") + ylab("Population difference between sources") + 
  facet_grid(vars(Country), vars(Year))
```

Fit & prediction `2020/21`:

```{r include=FALSE}
data <- bind_rows(
  
  hmd_pop_age_agg %>% 
    filter(Year >= 2015), 
  
  es_pop_age_new
  
) %>% 
  mutate(Source = "HMD") %>% 
  bind_rows(pop_age_2020 %>% 
              mutate(Source = "Estimate"))
```

```{r}
fn_age_plot <- function(country) {
  data %>% 
    filter(Country == country) %>% 
    ggplot(data = .,
           aes(x = Year, y = Population / 1000000, 
               color = factor(Source))) +
    geom_point() +
    geom_smooth(data = . %>% 
                  filter(Source == "HMD" & Year < 2020),
                colour = "grey40", size = 0.5,
                method = "lm", 
                se = FALSE) +
    # geom_smooth(data = . %>% 
    #               filter(Source == "HMD" & Year < 2020),
    #             method = "lm", 
    #             # formula = y ~ poly(x, 2, raw = TRUE), 
    #             formula = y ~ splines::bs(x, df = 3),
    #             # method = "gam", 
    #             #   formula = y ~ s(x), 
    #             se = FALSE) + 
    scale_y_continuous(labels = comma) +
    theme_bw() +
    scale_colour_discrete(hue_pal()(2), name = "Source") +
    xlab("Year") + ylab("Population [millions]") + 
    facet_wrap(vars(Age_cat), scales = "free_y")
}
```

## Spain

```{r echo=FALSE, warning=FALSE}
fn_age_plot("Spain")
```

## Sweden 

```{r echo=FALSE, warning=FALSE}
fn_age_plot("Sweden")
```

## Switzerland

```{r echo=FALSE, warning=FALSE}
fn_age_plot("Switzerland")
```

<!-- ----------------------------------------------------- -->

# Combined analysis datasets

## Population - predicted figures

Merging HMD until 2019 and predictions for 2020 & 2021.  

```{r}
pop_age_sex_exp <- bind_rows(
  
  hmd_pop_age_agg %>% 
    filter(Year < 2020), 
  
  pop_age_2020 
) %>% 
  arrange(Country, Year, Age_cat) %>% 
  rename(Population_exp = Population)

pop_yearly_exp <- pop_age_sex_exp %>% 
  group_by(Country, Year) %>% 
  summarise(Population_exp = sum(Population_exp)) %>% 
  ungroup()
```

## Population - observed figures

Merging HMD until 2021 for CHE & SWE with INE data for ESP for 2020 & 2021.  

```{r}
pop_age_sex_obs <- bind_rows(
  
  hmd_pop_age_agg, 
  
  es_pop_age_new 
) %>% 
  arrange(Country, Year, Age_cat) %>% 
  rename(Population_obs = Population)

pop_yearly_obs <- pop_age_sex_obs %>% 
  group_by(Country, Year) %>% 
  summarise(Population_obs = sum(Population_obs)) %>% 
  ungroup()
```

## Monthly deaths with yearly pops

```{r}
deaths_monthly <- bind_rows(
  
  ### ESP - starting at 1908; see issue #81
  es_deaths_month %>% 
    filter(Year >= 1908),
  
  # NAs for second half of year
  es_deaths_month %>% 
    filter(Year == 2020 & Month > 6) %>% 
    mutate(Year = as.integer(2021), 
           Deaths = NA_integer_),
  
  ### SWE
  se_deaths_month,
  
  se_deaths_month %>% 
    filter(Year == 2020 & Month > 6) %>% 
    mutate(Year = as.integer(2021), 
           Deaths = NA_integer_,
           Date = Date + years(1)),
  
  ch_deaths_month,
  
  ch_deaths_month %>% 
    filter(Year == 2020 & Month > 6) %>% 
    mutate(Year = as.integer(2021), 
           Deaths = NA_integer_,
           Date = Date + years(1))
  
) %>% 
  left_join(pop_yearly_obs) %>% 
  left_join(pop_yearly_exp) %>% 
  arrange(Country, Year, Month) %>% 
  mutate(si_one = sin(2*pi*Month/12),
         si_two = sin(4*pi*Month/12),
         co_one = cos(2*pi*Month/12),
         co_two = cos(4*pi*Month/12)) 

write_rds(deaths_monthly, "data/deaths_monthly.Rds")
```

```{r echo=FALSE}
skim(deaths_monthly)
```

## Yearly deaths by sex and age group with pops

Spain age&sex distribution in HMD is [available from later](https://github.com/RPanczak/ISPM_excess-mortality/issues/81)!

```{r}
deaths_yearly_age_sex <- bind_rows(
  
  hmd_deaths_age_sex,
  
  # SWE
  # 2020 is already in HMD for SWE 
  # 2021 is sadly missing for SWE 
  
  #ESP
  es_deaths_age_2019_2020,
  es_deaths_age_2021,
  
  # CHE
  # 2020 is already in HMD for CHE 
  ch_deaths_age_sex_2021
  
) %>% 
  mutate(
    # Age_cat = cut(Age, 
    # breaks = c(0, 1, 5, 20, 40, 60, 80, 112),
    # # labels = c("<1", "1-4", "5-19", "20-39", "40-59", "60-79","80+"), 
    # right = FALSE), 
    Age_cat = cut(Age, 
                  breaks = c(seq(0, 90, 10), 120), 
                  include.lowest = TRUE, right = FALSE)
  ) %>%
  group_by(Country, Year, Age_cat) %>% 
  summarise(Deaths = sum(Deaths)) %>% 
  ungroup() %>% 
  left_join(pop_age_sex_obs) %>% 
  left_join(pop_age_sex_exp) 

# labels=c("0-9","10-19","20-29","30-39","40-49", "50-59","60-69","70-79","80-89","90+")

write_rds(deaths_yearly_age_sex, "data/deaths_yearly_age_sex.Rds")
```

```{r echo=FALSE}
skim(deaths_yearly_age_sex)
```

<!-- ----------------------------------------------------- -->

# Computing Environment

`r mu$session()`