---
title: "Excess mortality: data preparation"
author: "Radek Panczak"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
editor_options: 
  chunk_output_type: console
---


```{r setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

options(scipen=999)
set.seed(12345)

library(pacman)
p_load(knitr, rmdformats, 
       tidyverse, scales, readxl, haven, janitor, kableExtra)

options(max.print = "75")
opts_chunk$set(cache = FALSE,
               prompt = FALSE,
               tidy = FALSE,
               comment = NA,
               message = FALSE,
               warning = FALSE)
opts_knit$set(width = 75)

mu <- Hmisc::markupSpecs$html
```

<!-- ----------------------------------------------------- -->

# HMD

[The Human Mortality Database](https://www.mortality.org/) data. 

```{r eval=FALSE}
p_load(HMDHFDplus)

password <- readr::read_file("secrets/HMD.txt")

# pop
hmd_ch_pop_raw <- readHMDweb(CNTRY = "CHE", item ="Population" , 
                             username = "r.panczak@gmail.com", password = password,
                             fixup = FALSE) %>% 
  as_tibble()

write_rds(hmd_ch_pop_raw, "data-raw/mortality_org/hmd_ch_pop_raw.Rds")
write_dta(hmd_ch_pop_raw, "data-raw/mortality_org/hmd_ch_pop_raw.dta")

# deaths
hmd_ch_deaths_raw <- readHMDweb(CNTRY = "CHE", item = "Deaths_1x1",
                                username = "r.panczak@gmail.com", password = password,
                                fixup = FALSE) %>% 
  as_tibble()

write_rds(hmd_ch_deaths_raw, "data-raw/mortality_org/hmd_ch_deaths_raw.Rds")
write_dta(hmd_ch_deaths_raw, "data-raw/mortality_org/hmd_ch_deaths_raw.dta")

hmd_ch_deaths <- hmd_ch_deaths_raw %>% 
  select(Year, Total) %>% 
  rename(year = Year) %>% 
  group_by(year) %>% 
  summarise(deaths_year = sum(Total)) %>% 
  as_tibble()

write_rds(hmd_ch_deaths, "data/mortality_org/hmd_ch_deaths.Rds")
write_dta(hmd_ch_deaths, "data/mortality_org/hmd_ch_deaths.dta")
p_unload(HMDHFDplus)
```

```{r include=FALSE}
hmd_ch_deaths_raw <- read_rds("data-raw/mortality_org/hmd_ch_deaths_raw.Rds")
hmd_ch_deaths <- read_rds("data/mortality_org/hmd_ch_deaths.Rds")
```

Available only yearly but for one year age bands & separate by sex. Time from `r min(hmd_ch_deaths_raw$Year)` to `r max(hmd_ch_deaths_raw$Year)`.

Data is in yearly resolution. 

```{r}
ggplot(hmd_ch_deaths, aes(x = year)) +
  geom_line(aes(y = deaths_year)) +
  theme_bw() +
  xlab("") + ylab("Yearly number of deaths")
```

<!-- ----------------------------------------------------- -->

# BfS 1

Data from [Todesfälle nach Monat und Sterblichkeit seit 1803](https://www.bfs.admin.ch/bfs/de/home/statistiken/kataloge-datenbanken/daten.assetdetail.14387168.html).

With update from @Kaspar covering missing months between 1886 and 1900.

```{r message=FALSE, warning=FALSE}
bfs_web_raw <- read_xlsx("data-raw/BfS/px-x-0102020206_111_20210223-102443.xlsx", 
                         skip = 2, na = "...") %>% 
  remove_empty() %>% 
  clean_names() %>% 
  rename(year = x1) %>% 
  filter(!is.na(todesfalle_total)) %>% 
  select(year:todesfalle_im_dezember) %>% 
  mutate(year = as.numeric(year))

bfs_web_deaths_yearly <- bfs_web_raw %>% 
  select(year, todesfalle_total) %>% 
  rename(deaths_year = todesfalle_total) 

p1 <- bfs_web_raw %>% 
  select(year, todesfalle_total) %>% 
  rename(deaths_year = todesfalle_total) %>% 
  filter(year < 1877) %>% 
  uncount(12) %>% 
  group_by(year) %>% 
  mutate(month_num = row_number()) %>% 
  ungroup() %>% 
  mutate(deaths_month = NA,
         type = "yearly")

bfs_2_web_raw <- read_xlsx("data-raw/BfS/Todesf_n_Monat_seit_1877_DEM_2.xlsx") %>% 
  remove_empty() %>% 
  clean_names() %>% 
  rename(year = jahr) 

p2 <- bfs_2_web_raw %>% 
  pivot_longer(!year, names_to = "month_num", values_to = "deaths_month") %>% 
  mutate(month_num = as.numeric(str_replace(month_num, fixed("m"), "")))  %>% 
  mutate(date = anytime::anydate(paste0(year, "-", month_num))) %>% 
  relocate(year, month_num, date) %>% 
  mutate(type = "monthly")

bfs_web_deaths_monthly <- p1 %>% 
  bind_rows(p2) %>% 
  mutate(date = anytime::anydate(paste0(year, "-", month_num))) %>% 
  relocate(year, month_num, date) 

write_rds(bfs_web_deaths_monthly, "data/BfS/bfs_web_deaths_monthly.Rds")
write_dta(bfs_web_deaths_monthly, "data/BfS/bfs_web_deaths_monthly.dta")
```

Yearly deaths for period 1803 to 1876; monthly deaths thereafter.

```{r}
ggplot(data = filter(bfs_web_deaths_monthly, 
                       year < 1877), 
       aes(x = date)) +
  geom_line(aes(y = deaths_year)) +
  geom_line(data = filter(bfs_web_deaths_monthly, 
                            year >= 1877), 
            aes(y = deaths_month)) +
  facet_grid(vars(type), scales = "free_y") +
  theme_bw() +
  xlab("") + ylab("Monthly/yearly number of deaths")
```

Note different `y` scales! 

<!-- ----------------------------------------------------- -->

# Russian Flu

```{r message=FALSE, warning=FALSE}
extra_1889 <- read_xlsx("data-raw/Kaspar/1889 CH.xlsx") %>% 
  remove_empty() %>% 
  clean_names() %>% 
  rename(year = x1) %>% 
  pivot_longer(-year, values_to = "deaths_month") %>% 
  mutate(month_num = rep(1:12, length(unique(year))),
         type = "monthly",
         date = anytime::anydate(paste0(year, "-", month_num))) 
```

Extra monthly dataset covering 'Russian Flu' between `r min(extra_1889$year)` and `r max(extra_1889$year)`. 

```{r}
ggplot(extra_1889) +
  geom_line(aes(x = date, y = deaths_month)) +
  theme_bw() +
  xlab("") + ylab("Monthly number of deaths")
```

<!-- ----------------------------------------------------- -->

# Comparison

## Yearly vs. monthly figures

```{r}
bfs_web_deaths_monthly %>% 
  filter(!is.na(deaths_month)) %>% 
  group_by(year) %>% 
  summarize(deaths_summed = sum(deaths_month)) %>% 
  ungroup() %>% 
  left_join(bfs_web_deaths_yearly) %>% 
  filter(deaths_year != deaths_summed) %>% 
  nrow()
```

There are no years where summed monthly data differ from 'official' yearly sums.

## Russian Flu vs BfS

```{r}
compare <- extra_1889 %>% 
  select(date, deaths_month) %>% 
  rename(deaths_month_russian = deaths_month) %>% 
  left_join(bfs_web_deaths_monthly %>% 
              select(date, deaths_month) %>% 
              rename(deaths_month_bfs = deaths_month)) %>% 
  mutate(diff = deaths_month_bfs - deaths_month_russian)
```

Summary of differences between BfS figures and HMD

```{r}
summary(compare$diff)
```

Timeline of differences between BfS figures and HMD

```{r}
ggplot(compare, aes(x = date, y = diff)) +
  geom_line() +
  theme_bw() +
  xlab("") + ylab("Yearly difference in number of deaths")
```

## BfS vs HMD

```{r}
compare <- left_join(
  bfs_web_deaths_yearly %>% 
    rename(deaths_year_bfs = deaths_year), 
  hmd_ch_deaths %>% 
    rename(deaths_year_hmd = deaths_year)) %>% 
  mutate(diff = deaths_year_bfs - deaths_year_hmd)
```

Summary of differences between BfS figures and HMD

```{r}
summary(compare$diff)
```

Timeline of differences between BfS figures and HMD

```{r}
ggplot(compare, aes(x = year, y = diff)) +
  geom_line() +
  theme_bw() +
  xlab("") + ylab("Yearly difference in number of deaths")
```

<!-- ----------------------------------------------------- -->

# Computing Environment

`r mu$session()`