---
title: "Excess mortality"
subtitle: "Weekly data"
author: "Radek Panczak"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
---

<!-- ----------------------------------------------------- -->

```{r setup, include = FALSE}
set.seed(12345)
options(scipen = 999)
options(max.print = "75")

library(pacman)
p_load(tidyverse, readxl, scales, janitor, kableExtra,
       lubridate, aweek, wktmo)
```

```{r knit-setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_chunk$set(cache = FALSE,
                      prompt = FALSE,
                      tidy = FALSE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE)

knitr::opts_knit$set(width = 75)

mu <- Hmisc::markupSpecs$html
```

<!-- ----------------------------------------------------- -->

# STMF weekly data series

Short-term Mortality Fluctuations weekly dataset covering 2000-2020 period.  

```{r eval=FALSE}
download.file(url = "https://www.mortality.org/Public/STMF/Outputs/stmf.csv",
              destfile = "data-raw/mortality_org/stmf_week.csv",
              method = "curl")

zip(zipfile = "data-raw/mortality_org/stmf_week", 
    files = "data-raw/mortality_org/stmf_week.csv")

file.remove("data-raw/mortality_org/stmf_week.csv")

download.file(url = "https://www.mortality.org/Public/STMF_DOC/STMFmetadata.pdf",
              destfile = "data-raw/mortality_org/STMFmetadata.pdf",
              method = "curl")

download.file(url = "https://www.mortality.org/Public/STMF_DOC/STMFNote.pdf",
              destfile = "data-raw/mortality_org/STMFNote.pdf",
              method = "curl")
```

## Weekly

```{r}
stmf_week <- read_csv("data-raw/mortality_org/stmf_week.zip", 
                      col_types = cols(Year = col_integer(), 
                                       Week = col_integer(), 
                                       Split = col_integer(), 
                                       SplitSex = col_integer(), 
                                       Forecast = col_integer()), 
                      skip = 1) %>% 
  filter(CountryCode %in% c("CHE", "ESP", "SWE")) %>% 
  filter(Sex == "b") %>% 
  select(CountryCode, Year, Week, DTotal) %>% 
  rename(Deaths = DTotal) %>%  
  mutate(Week2 = get_aweek(week = Week, year = Year, 
                           start = "Monday", week_start = "Monday"),
         Week_start = get_date(week = Week, year = Year, 
                               start = "Monday"),
         Deaths = as.integer(Deaths)) %>% 
  relocate(Week2, Week_start, .after = Week) %>% 
  mutate(Country = case_when(
    CountryCode == "CHE" ~ "Switzerland",
    CountryCode == "ESP" ~ "Spain",
    CountryCode == "SWE" ~ "Sweden"
  )) %>% select(-CountryCode) %>% relocate(Country)

write_rds(stmf_week, "data/mortality_org/stmf_week.Rds")
```

```{r echo=FALSE}
stmf_week %>% 
  filter(Year >= 2019) %>% 
  ggplot(aes(x = Week_start)) +
  geom_line(aes(y = Deaths, group = Country, color = Country)) +
  theme_bw() +
  xlab("") + ylab("Weekly number of deaths")
```

Switzerland & Sweden have slightly more data:  

```{r}
stmf_week %>% 
  filter(Year == 2021) %>% 
  group_by(Country) %>% 
  summarize(Max_week = max(Week)) 
```

## Monthly

Weekly data converted to monthly using [weekToMonth](https://rdrr.io/cran/wktmo/man/weekToMonth.html) function from `wktmo` package.  

```{r}
stmf_month <- bind_rows(
  
  weekToMonth(stmf_week[stmf_week$Country == "Switzerland", ]$Deaths, 
              datStart = "03-01-2000", wkMethod = "ISO") %>% 
    as_tibble() %>% 
    mutate(Country = "Switzerland"),
  
  weekToMonth(stmf_week[stmf_week$Country == "Spain", ]$Deaths, 
              datStart = "03-01-2000", wkMethod = "ISO") %>% 
    as_tibble() %>% 
    mutate(Country = "Spain"),
  
  weekToMonth(stmf_week[stmf_week$Country == "Sweden", ]$Deaths, 
              datStart = "03-01-2000", wkMethod = "ISO") %>% 
    as_tibble() %>% 
    mutate(Country = "Sweden") 
  
) %>% 
  separate(yearMonth, c("Year", "Month"), convert = TRUE) %>% 
  rename(Deaths = value) %>% 
  mutate(Date = ymd(paste0(Year, "-", Month, "-01")),
         Deaths = as.integer(Deaths)) %>% 
  relocate(Country, Year, Month, Date)

write_rds(stmf_month, "data/mortality_org/stmf_month.Rds")
```

```{r echo=FALSE}
stmf_month %>% 
  filter(
    (Date <= as.Date("2021-08-01") & Country == "Spain") |
      (Date <= as.Date("2021-09-01") & Country != "Spain") 
  ) %>%
  filter(Year >= 2019) %>% 
  ggplot(aes(x = Date)) +
  geom_line(aes(y = Deaths, group = Country, color = Country)) +
  theme_bw() +
  xlab("") + ylab("Monthly number of deaths")
```

*Note: Only showing data till Aug 2021 for Spain (data is available only till week 38 which ends on `2021-09-26`). Data till Sept for Switzerland and Sweden.*  
## Yearly totals

Conversions are not perfect!  

```{r}
differences <- left_join(
  stmf_week %>% 
    filter(Year >= 2019) %>% 
    group_by(Country, Year) %>% 
    summarise(Weekly = round(sum(Deaths))) %>% 
    ungroup(),
  
  stmf_month %>% 
    filter(Year >= 2019) %>% 
    group_by(Country, Year) %>% 
    summarise(Monthly = round(sum(Deaths))) %>% 
    ungroup()
) %>% 
  mutate(Difference = Weekly - Monthly,
         Perc_weekly = percent((Weekly - Monthly) / Weekly, accuracy = 0.01))
```

Example of 2019-21 yearly totals derived from both original weekly & transformed monthly data illustrate this problem:

```{r echo=FALSE}
differences %>% 
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>% 
  column_spec(1, italic = TRUE) %>%
  row_spec(1, italic = FALSE)  %>% 
  column_spec(c(5,6), 
              color = ifelse(differences$Difference > 0, 
                             "#EF8A62", "#67A9CF"))
```

```{r include=FALSE}
rm(differences)
```

Showing small imperfections of the `week -> month` transition! 2020 seems to be always a bit in `+` and 2021 in `-`.  

<!-- ----------------------------------------------------- -->

# BfS weekly data series for CHE

*Mortality, causes of death* chapter. [Overview of data](https://www.bfs.admin.ch/bfs/en/home/statistics/health/state-health/mortality-causes-death.html).  

## Weekly number of deaths, 2010-2019

```{r eval=FALSE}
download.file(url = "https://www.bfs.admin.ch/bfsstatic/dam/assets/12607335/master",
              destfile = "data-raw/BfS/ts-e-14.03.04.03-wr_ZR.csv",
              method = "curl")
```

```{r include=FALSE}
week_2010_2019 <- read_delim("data-raw/BfS/ts-e-14.03.04.03-wr_ZR.csv", ";",
                             escape_double = FALSE, 
                             col_types = cols(CY = col_integer(), 
                                              Week = col_integer(), 
                                              Ending = col_date(format = "%d.%m.%Y"), 
                                              NumberOfDeaths = col_integer(), 
                                              Expected = col_integer(), 
                                              lowerB = col_integer(), 
                                              upperB = col_integer(), 
                                              Excess = col_integer()), 
                             trim_ws = TRUE) %>% 
  rename(Year = CY) %>% 
  filter(!is.na(Year))
```

Last five years with two age groups

```{r echo=FALSE}
week_2010_2019 %>% 
  filter(Year >= 2015) %>% 
  ggplot(aes(x = Ending)) +
  # geom_vline(aes(xintercept = as.Date("2020-02-24"))) +
  geom_ribbon(aes(ymin = lowerB, ymax = upperB, fill = Age), alpha = 0.1) +
  geom_line(aes(y = Expected, group = Age), colour = "gray60") + 
  geom_line(aes(y = NumberOfDeaths, colour = Age)) + 
  theme_minimal()
```

## Weekly number of deaths, 2020-2021

[Source](https://www.bfs.admin.ch/bfs/en/home/statistics/catalogues-databases/tables.assetdetail.19564278.html).  

```{r eval=FALSE}
download.file(url = "https://www.bfs.admin.ch/bfsstatic/dam/assets/19564278/master",
              destfile = "data-raw/BfS/ts-e-14.03.04.03-wr.csv",
              method = "curl")

zip(zipfile = "data-raw/BfS/ts-e-14.03.04.03-wr.zip", 
    files = "data-raw/BfS/ts-e-14.03.04.03-wr.csv")

file.remove("data-raw/BfS/ts-e-14.03.04.03-wr.csv")
```

```{r include=FALSE}
week_2020_2021 <- read_delim("data-raw/BfS/ts-e-14.03.04.03-wr.zip", 
                             ";", escape_double = FALSE, 
                             col_types = cols(Year = col_integer(), 
                                              Week = col_integer(), 
                                              Ending = col_date(format = "%d.%m.%Y"), 
                                              NoDeaths_EP = col_double(), 
                                              Expected = col_double(), 
                                              LowerB = col_double(), 
                                              UpperB = col_double(), 
                                              Diff = col_double()), 
                             trim_ws = TRUE, na = c(".")) %>% 
  rename(NumberOfDeaths = NoDeaths_EP,
         lowerB = LowerB, upperB = UpperB,
         Excess = Diff) %>% 
  filter(!is.na(Year))
```

```{r echo=FALSE}
week_2020_2021 %>% 
  filter(Ending < Sys.Date()) %>% 
  ggplot(aes(x = Ending)) +
  # geom_vline(aes(xintercept = as.Date("2020-02-24"))) +
  geom_ribbon(aes(ymin = lowerB, ymax = upperB, fill = Age), alpha = 0.1) +
  geom_line(aes(y = Expected, group = Age), colour = "gray60") + 
  geom_line(aes(y = NumberOfDeaths, colour = Age)) + 
  theme_minimal()
```

```{r echo=FALSE,eval=FALSE}
# projected goes on; with a funny dip at the end?
week_2020_2021 %>% 
  filter(Ending >= as.Date("2020-11-01")) %>% 
  ggplot(aes(x = Ending)) +
  # geom_vline(aes(xintercept = as.Date("2020-02-24"))) +
  geom_ribbon(aes(ymin = lowerB, ymax = upperB, fill = Age), alpha = 0.1) +
  geom_line(aes(y = Expected, group = Age), colour = "gray60") + 
  geom_line(aes(y = NumberOfDeaths, colour = Age)) + 
  theme_minimal()
```

```{r include=FALSE}
week_2010_2021 <- week_2010_2019 %>% 
  bind_rows(week_2020_2021) %>% 
  write_rds("data/BfS/deaths_week_2010-2021.Rds")

rm(week_2010_2019, week_2020_2021)
```

## Yearly totals of excess deaths

**2021 calculated untill week 25 (`2021-06-27`)**  

```{r}
excess_2010_2020 <- week_2010_2021 %>% 
  filter(Ending <= as.Date("2021-06-27")) %>% 
  mutate(Excess_expected = NumberOfDeaths - Expected) %>% 
  # mutate(Excess_upperB = NumberOfDeaths - upperB) %>% 
  # group_by(Year, Age) %>% 
  group_by(Year) %>% 
  summarise(Excess_expected = sum(Excess_expected)) %>% 
  # summarise(Excess_upperB = sum(Excess_upperB)) %>% 
  ungroup()

write_rds(excess_2010_2020, "data/BfS/excess_2010_2020.Rds")
```

```{r echo=FALSE}
ggplot(excess_2010_2020) + 
  geom_col(aes(x = factor(Year), y = Excess_expected)) + 
  theme_minimal() + xlab("") + ylab("Excess deaths")
```

<!-- ----------------------------------------------------- -->

# Comparing BfS to STMF {.tabset}

Data from 2020 onwards, weekly resolution.  

```{r}
bfs_stmf <- left_join(
  
  week_2010_2021 %>% 
    filter(Ending <= as.Date("2021-05-23")) %>% 
    group_by(Year, Week) %>%
    summarise(NumberOfDeaths = sum(NumberOfDeaths)) %>% 
    ungroup(),
  
  stmf_week %>% 
    filter(Country == "Switzerland")
  
) %>% 
  mutate(diff = NumberOfDeaths - Deaths)
```

Some end of the year corrections (?):

## Absolute 

```{r echo=FALSE}
bfs_stmf %>% 
  filter(Year >= 2015) %>% 
  ggplot(aes(x = Week_start, y = diff)) + 
  geom_hline(yintercept = 0, color = "darkorchid") +
  geom_vline(xintercept = as.Date("2021-05-01"), color = "orange") +
  geom_col() + 
  theme_bw() +
  xlab("") + ylab("Weekly absolute difference in number of deaths")
```

## Percentages 

```{r echo=FALSE}
bfs_stmf %>% 
  filter(Year >= 2015) %>% 
  ggplot(aes(x = Week_start, y = diff / NumberOfDeaths)) + 
  geom_hline(yintercept = 0, color = "darkorchid") +
  geom_vline(xintercept = as.Date("2021-05-01"), color = "orange") +
  geom_col() + 
  theme_bw() +
  xlab("") + ylab("Weekly % difference in number of deaths") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))
```

<!-- ----------------------------------------------------- -->

# Computing Environment

`r mu$session()`