---
title: "Excess mortality: weekly data & `surveillance` package"
author: "Radek Panczak"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
editor_options: 
  chunk_output_type: console
---


```{r setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

options(scipen=999)
set.seed(12345)

library(pacman)
p_load(tidyverse, scales, kableExtra, surveillance)

options(max.print = "75")
knitr::opts_chunk$set(cache = FALSE,
                      prompt = FALSE,
                      tidy = FALSE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE)
knitr::opts_knit$set(width = 75)

mu <- Hmisc::markupSpecs$html
```

<!-- ----------------------------------------------------- -->

# Weekly number of deaths, 2010-2019

```{r include=FALSE}
weekly_2010_2019 <- read_rds("data/BfS/weekly_2010-2019.Rds") %>% 
  mutate(pandemic = 0)
weekly_2020_2021 <- read_rds("data/BfS/weekly_2020-2021.Rds") %>% 
  mutate(pandemic = 1)
weekly_2010_2021 <- weekly_2010_2019 %>% 
  bind_rows(weekly_2020_2021) %>% 
  filter(!is.na(NumberOfDeaths))
```

Last decade(ish)

```{r echo=FALSE}
weekly_2010_2021 %>% 
  # filter(Year >= 2015) %>% 
  ggplot(aes(x = Ending)) +
  geom_ribbon(aes(ymin = lowerB, ymax = upperB, fill = Age), alpha = 0.1) +
  geom_line(aes(y = Expected, group = Age), colour = "gray60") + 
  geom_line(aes(y = NumberOfDeaths, colour = Age)) + 
  theme_minimal() + 
  scale_y_continuous(labels = comma) +
  ylab("") + xlab("Number of deaths")
```

# Data structure

Package works with `srs` - *surveillance time series* objects. Creating one for 65 and older

```{r}
deaths_65 <- weekly_2010_2021 %>% 
  filter(Age == "65+") %>% 
  select(-Age)

deaths_65_sts <- sts(deaths_65$NumberOfDeaths, 
                     start = c(2010, 1), 
                     frequency = 52)

# deaths_65_sts <- new("sts", 
#                      epoch = as.numeric(deaths_65$Ending),
#                      observed = deaths_65$NumberOfDeaths, 
#                      epochAsDate = TRUE)
```

Also selecting indices of observations in the pandemic

```{r}
in2020 <- which(deaths_65$Ending >= as.Date("2020-02-01"))
```

# Three methods 

##  EARS method {.tabset}

With three sub methods

### C1

EARS C1-MILD method 

```{r}
control <- list(range = in2020, method = "C1", alpha = 0.05)
surv <- earsC(deaths_65_sts, control = control)
```

```{r echo=FALSE}
plot(surv)
```

### C2

EARS C2-MEDIUM method

```{r}
control <- list(range = in2020, method = "C2", alpha = 0.05)
surv <- earsC(deaths_65_sts, control = control)
```

```{r echo=FALSE}
plot(surv)
```

### C3

EARS C3-HIGH method

```{r}
control <- list(range = in2020, method = "C3", alpha = 0.05)
surv <- earsC(deaths_65_sts, control = control)
```

```{r echo=FALSE}
plot(surv)
```

# `farringtonFlexible` original method

```{r}
control1 <- list(range = in2020, 
                 noPeriods = 1, b = 4, w = 3,
                 weightsThreshold = 1, 
                 pastWeeksNotIncluded = 3,
                 pThresholdTrend = 0.05, 
                 thresholdMethod = "delta")

farrington <- farringtonFlexible(deaths_65_sts, control1)
```

```{r echo=FALSE}
plot(farrington)
```

# `farringtonFlexible` improved method

```{r}
control2 <- list(range = in2020, 
                 noPeriods = 10, b = 4, w = 3,
                 weightsThreshold = 2.58, 
                 pastWeeksNotIncluded = 26,
                 pThresholdTrend = 1, 
                 thresholdMethod = "nbPlugin")

noufaily <- farringtonFlexible(deaths_65_sts, control2)
```

```{r echo=FALSE}
plot(noufaily)
```

<!-- ----------------------------------------------------- -->

# Computing Environment

`r mu$session()`