---
title: "Excess mortality"
subtitle: "Comparison of monthly and weekly data use"
author: "Radek Panczak"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  code_folding: show
self_contained: true
highlight: pygments
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
---

<!-- ----------------------------------------------------- -->

```{r setup, include = FALSE}
set.seed(12345)
options(scipen = 999)
options(max.print = "75")

library(pacman)
p_load(tidyverse, magrittr, scales, aweek, 
       cmdstanr, posterior)
```

```{r knit-setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_chunk$set(cache = FALSE,
                      prompt = FALSE,
                      tidy = FALSE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE)
knitr::opts_knit$set(width = 75)

mu <- Hmisc::markupSpecs$html
```

<!-- ----------------------------------------------------- -->

# Data & set up

## Misc

```{r}
## Counrtry 
COUNTRIES = c("Switzerland","Spain","Sweden")

## Pandemics & subsequent post-pandemic years to exclude
pandemic <- 2020

## Funcs
source("R/fn_global_serfling_nb_week_cmdstan.R")

## Starting year
START <- 2000
```

## Monthly data 

```{r}
## Population denominator
pops <- read_rds("data/deaths_monthly.Rds") %>% 
  filter(Month == 1) %>% 
  select(Year, Country, Population_obs) 

stmf_month <- read_rds("data/mortality_org/stmf_month.Rds") %>% 
  left_join(pops) %>% 
  mutate(si_one = sin(2*pi*Month/12),
         si_two = sin(4*pi*Month/12),
         co_one = cos(2*pi*Month/12),
         co_two = cos(4*pi*Month/12)) 
```

## Weekly data

Recalculating sin & cos data:  

```{r}
stmf_week <- read_rds("data/mortality_org/stmf_week.Rds") %>% 
  left_join(pops) %>% 
  mutate(
    si_one = sin(2*pi*Week/(365.25/7)),
    si_two = sin(4*pi*Week/((365.25/2)/7)),
    co_one = cos(2*pi*Week/(365.25/7)),
    co_two = cos(4*pi*Week/((365.25/2)/7))
  )

rm(pops)
```

Getting half of the year in 2021 - data kept until **Week 26**. That means it's *not strictly speaking half a year* since week 26 runs four days into July.  

```{r}
stmf_week %<>% 
  filter(! (Year == 2021 & Week > 26))
```

Problem of 53! Some years have one more week. Will be resolved later.  

```{r echo=FALSE}
stmf_week %>% 
  filter(Week == 53) %>% 
  group_by(Year) %>% 
  summarise(Year = first(Year))
```

# Monthly results 

```{r}
monthly_results <- tibble()
monthly_results_year <- tibble()

for(COUNTRY in COUNTRIES) {
  
  ## Data 
  deaths_monthly <- stmf_month %>% 
    filter(Country == COUNTRY)
  
  ## Loop
  for (YEAR in (START+5):2021) {
    
    print(paste("Country:", COUNTRY, "Year:", YEAR, "Monthly loop."))
    
    m_glo <- fn_global_serfling_nb_cmdstan(YEAR, 
                                           deaths_monthly, 
                                           pandemic_years = pandemic, 
                                           pop = "obs",
                                           version = "last_5")
    
    
    monthly_results <- m_glo$pred_total_deaths %>% 
      dplyr::select(-starts_with("excess_year")) %>% 
      dplyr::mutate(Model = "Monthly") %>% 
      bind_rows(., monthly_results)
    
    monthly_results_year <- m_glo$pred_total_deaths %>% 
      filter(row_number() == 1) %>% 
      select(Country, Year, starts_with("excess_year")) %>% 
      dplyr::mutate(Model = "Monthly") %>% 
      bind_rows(., monthly_results_year)
    
    rm(m_glo)
    gc()
  }
}
```

# Weekly results 

Models estimated using **52 weeks only**!  

```{r}
weekly_results <- tibble()
weekly_results_year <- tibble()

for(COUNTRY in COUNTRIES) {
  
  ## Data 
  deaths_weekly <- stmf_week %>% 
    filter(Country == COUNTRY) %>% 
    filter(Week < 53)
  
  ## Loop
  for (YEAR in (START+5):2021) {
    
    print(paste("Country:", COUNTRY, "Year:", YEAR, "Monthly loop."))
    
    m_glo <- fn_global_serfling_nb_week_cmdstan(YEAR, 
                                                deaths_weekly, 
                                                pandemic_years = pandemic, 
                                                pop = "obs")
    
    
    weekly_results <- m_glo$pred_total_deaths %>% 
      dplyr::select(-starts_with("excess_year")) %>% 
      dplyr::mutate(Model = "Weekly") %>% 
      bind_rows(., weekly_results)
    
    weekly_results_year <- m_glo$pred_total_deaths %>% 
      filter(row_number() == 1) %>% 
      select(Country, Year, starts_with("excess_year")) %>% 
      dplyr::mutate(Model = "Weekly") %>% 
      bind_rows(., weekly_results_year)
    
    rm(m_glo)
    gc()
  }
}
```

```{r include=FALSE}
rm(deaths_monthly, deaths_weekly, COUNTRIES, pandemic, START); gc()
```

```{r echo=FALSE}
weekly_results_53 <- weekly_results %>% 
  bind_rows(
    weekly_results %>% 
      filter(Year %in% c(2009, 2015, 2020)) %>% 
      filter(Week == 52) %>% 
      mutate(Week = 53) %>% 
      select(Country, Year, Week, Population, pred, lower, upper, Model) %>% 
      left_join(
        
        stmf_week %>% 
          filter(Year %in% c(2009, 2015, 2020)) %>% 
          filter(Week == 53) %>% 
          select(Country, Year, Week, Deaths) 
        
      ) %>% 
      mutate(
        excess_week = Deaths - pred,
        excess_week_lower = Deaths - lower, 
        excess_week_upper = Deaths - upper
      )
  ) %>% 
  arrange(Country, Year, Week) %>% 
  mutate(Date = get_date(Week, Year)) %>% 
  relocate(Date, .after = Week)
```

# Results

## Example of one country, two years

```{r}
coeff <- 5

weekly_results_53 %>% 
  filter(Country == "Switzerland" & (Year == 2018 | Year == 2020)) %>% 
  ggplot() +
  geom_col(aes(x = Date, y = pred)) +
  geom_line(aes(x = Date, y = Deaths), colour = "red") +
  geom_line(data = monthly_results %>% 
              filter(Country == "Switzerland" & (Year == 2018 | Year == 2020)), 
            aes(x = Date + 15,
                y = pred / coeff),
            color = "black") +
  facet_grid(vars(Year), scales = "free_x") +
  scale_y_continuous(
    
    name = "Weekly counts",
    
    # Add a second axis and specify its features
    sec.axis = sec_axis(~.*coeff, name = "Monthly counts")
  ) + 
  theme_light()
```

## Differences of yearly totals of excess {.tabset}

Difference of weekly data summed to a year (with week 53 added) and yearly estimates from monthly data.  

```{r echo=FALSE}
differences <- monthly_results_year %>% 
  left_join(
    weekly_results_year_53 <- weekly_results_53 %>% 
      group_by(Country, Year) %>% 
      summarise(excess_year_weekly = sum(excess_week)) %>% 
      ungroup()
  ) %>% 
  mutate(absolute = excess_year - excess_year_weekly,
         relative = absolute / excess_year)
```

### Relative 

```{r}
ggplot(differences, aes(x = Year, y = relative)) +
  geom_col() +
  facet_grid(vars(Country))
```

### Absolute 

```{r}
ggplot(differences, aes(x = Year, y = absolute)) +
  geom_col() +
  facet_grid(vars(Country))
```



