---
title: "Excess mortality"
subtitle: "Alternative data sources"
author: "Radek Panczak"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
---

<!-- ----------------------------------------------------- -->

```{r setup, include = FALSE}
set.seed(12345)
options(scipen = 999)
options(max.print = "75")

library(pacman)
p_load(tidyverse, magrittr, janitor, scales,
       readxl, lubridate, aweek, wktmo,
       kableExtra)

import::from("pxR", "read.px")
```

```{r knit-setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_chunk$set(cache = FALSE,
                      prompt = FALSE,
                      tidy = FALSE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE)

knitr::opts_knit$set(width = 75)

mu <- Hmisc::markupSpecs$html
```

<!-- ----------------------------------------------------- -->

# Main data sources

## HMD yearly deaths with age structure

```{r}
hmd_deaths_age_sex <- read_rds("data/mortality_org/hmd_deaths_age_sex.Rds")
hmd_deaths_year <- read_rds("data/mortality_org/hmd_deaths_year.Rds")
hmd_pop_age_sex <- read_rds("data/mortality_org/hmd_pop_age_sex.Rds")
hmd_pop_year <- read_rds("data/mortality_org/hmd_pop_year.Rds")
```

## STMF weekly & monthly

```{r}
stmf_week <- read_rds("data/mortality_org/stmf_week.Rds")
stmf_month <- read_rds("data/mortality_org/stmf_month.Rds")
```

## BfS

```{r}
ch_deaths_month <- read_rds("data/BfS/ch_deaths_month.Rds")
```

<!-- ----------------------------------------------------- -->

# Switzerland 

## Comparison - deaths

### BfS vs HMD {.tabset}

Aggregating monthly deaths from BfS to year.  

```{r}
compare <- left_join(
  # ch_deaths_month_web %>% 
  ch_deaths_month %>%
    group_by(Year) %>% 
    summarise(deaths_year_bfs = sum(Deaths)), 
  hmd_deaths_year %>% 
    filter(Country == "Switzerland") %>% 
    rename(deaths_year_hmd = Deaths)) %>% 
  mutate(diff = deaths_year_bfs - deaths_year_hmd,
         perc = (deaths_year_bfs - deaths_year_hmd) / deaths_year_bfs)
```

Timeline of differences between yearly BfS and HMD figures. 

#### Absolute

```{r}
summary(compare$diff)
```

```{r echo=FALSE}
ggplot(compare, aes(x = Year, y = diff)) +
  geom_hline(yintercept = 0, colour = "lightblue") +
  geom_col() +
  theme_bw() +
  xlab("") + ylab("Yearly difference in number of deaths")
```

#### Percentage

```{r}
summary(compare$perc*100)
```

```{r echo=FALSE}
ggplot(compare, aes(x = Year, y = perc*100)) +
  geom_hline(yintercept = 0, colour = "lightblue") +
  geom_col() +
  theme_bw() +
  xlab("") + ylab("Yearly % difference in number of deaths")
```

### BfS vs STMF  {.tabset}

Comparing monthly figures.^[see data prep document for week <> month issues!]  

```{r}
compare <- inner_join(
  ch_deaths_month %>% 
    rename(deaths_bfs = Deaths),
  stmf_month %>% 
    filter(Country == "Switzerland") %>% 
    rename(deaths_stmf = Deaths) 
) %>% 
  mutate(diff = deaths_bfs - deaths_stmf,
         perc = (deaths_bfs - deaths_stmf) / deaths_bfs)
```

Summary of differences between monthly BfS and STMF figures. 

#### Absolute

```{r}
summary(compare$diff)
```

```{r echo=FALSE}
ggplot(compare, aes(x = Date, y = diff)) +
  geom_hline(yintercept = 0, colour = "lightblue") +
  geom_col() +
  theme_bw() +
  xlab("") + ylab("Absolute difference in number of deaths")
```

#### Percentage

```{r}
summary(compare$perc*100)
```

```{r echo=FALSE}
ggplot(compare, aes(x = Date, y = perc*100)) +
  geom_hline(yintercept = 0, colour = "lightblue") +
  geom_col() +
  theme_bw() +
  xlab("") + ylab("% difference in number of deaths")
```

```{r include=FALSE}
rm(compare)
```

## Yearly 2019-2020 deaths with age & sex structure 

Data from BfS website [Deaths by sex, citizenship, marital status and age](https://www.bfs.admin.ch/bfs/en/home/statistics/population.assetdetail.17664400.html).^[Or in custom tabs [here](https://www.pxweb.bfs.admin.ch/pxweb/en/px-x-0102020206_103/px-x-0102020206_103/px-x-0102020206_103.px/).]   

```{r}
ch_deaths_age_sex_2019_2020 <- read.px("data-raw/BfS/px-x-0102020206_103.px") %>% 
  as_tibble() %>% remove_empty() %>% clean_names() %>% 
  filter(zivilstand == "Zivilstand - Total",
         staatsangehorigkeit == "Staatsangehörigkeit - Total") %>% 
  select(-zivilstand, -staatsangehorigkeit) %>% 
  filter(alter != "Alter - Total") %>% 
  rename(Year = jahr,
         Total = value) %>% 
  mutate(Year = as.integer(as.character(Year)),
         geschlecht = as.character(geschlecht),
         Age = as.integer(word(alter, 1)),
         Total = as.integer(round(Total))) %>% 
  filter(Year >= 2019) %>% 
  select(-alter) %>% 
  mutate(geschlecht = if_else(geschlecht == "Geschlecht - Total", "Total", geschlecht)) %>% 
  mutate(geschlecht = if_else(geschlecht == "Mann", "Male", geschlecht)) %>% 
  mutate(geschlecht = if_else(geschlecht == "Frau", "Female", geschlecht)) %>% 
  pivot_wider(names_from = geschlecht, values_from = Total) %>% 
  mutate(Country = "Switzerland") %>% 
  relocate(Country, Year, Age, Female, Male, Total)
```

```{r message=FALSE, include=FALSE}
gc()
```

### Comparison to HMD results for 2019 & 2020. {.tabset}

```{r}
compare <- hmd_deaths_age_sex %>% 
  filter(Country == "Switzerland" & Year >= 2019) %>% 
  mutate(Age = ifelse(Age > 100, 100, Age)) %>% 
  select(-Female, -Male) %>% 
  group_by(across(c(-Total))) %>% 
  summarise(Total = sum(Total)) %>% 
  ungroup() %>% 
  rename(Total_hmd = Total) %>% 
  left_join(ch_deaths_age_sex_2019_2020 %>% 
              mutate(Age = ifelse(Age > 100, 100, Age)) %>% 
              select(-Female, -Male) %>% 
              group_by(across(c(-Total))) %>% 
              summarise(Total = sum(Total)) %>% 
              ungroup() %>% 
              rename(Total_bfs = Total) 
  ) %>% 
  mutate(diff = Total_bfs - Total_hmd,
         diff_p = ((Total_bfs - Total_hmd) / Total_hmd) *100)
```

#### Absolute difference

```{r}
summary(compare$diff)
```

```{r echo=FALSE}
ggplot(compare, aes(x = diff)) +
  geom_histogram() + 
  facet_grid(~Year)
```

```{r echo=FALSE}
ggplot(compare, aes(x = Age, y = diff)) +
  geom_col() + 
  facet_grid(~Year)
```

#### Percentage difference

```{r}
summary(compare$diff_p)
```

```{r echo=FALSE}
ggplot(compare, aes(x = diff_p)) +
  geom_histogram() + 
  facet_grid(~Year)
```

```{r echo=FALSE}
ggplot(compare, aes(x = Age, y = diff_p)) +
  geom_col() + 
  facet_grid(~Year)
```

```{r include=FALSE}
rm(compare)
```


## Yearly pops untill 2020

BfS data from [Demographic balance of the permanent resident population, 1861-2020](https://www.bfs.admin.ch/bfs/en/home/statistics/catalogues-databases/tables.assetdetail.18344368.html). Population of Jan and Dec averaged.^[Alternative source might be [here](https://www.bfs.admin.ch/bfs/en/home/statistics/catalogues-databases/data.assetdetail.19105850.html).]  

```{r}
ch_pop_year <- read_xlsx("data-raw/BfS/su-e-01.02.04.05.xlsx", 
                         na = "...", skip = 1) %>% 
  remove_empty(c("rows", "cols")) %>%
  clean_names() %>% 
  rename(Year = x1, pop_jan = population_on_2, pop_dec = population_on_11) %>%
  filter(!is.na(Year)) %>% 
  filter(!is.na(pop_jan)) %>% 
  mutate(Year = if_else(Year == "2010 7", "2010", Year),
         Year = as.integer(Year), 
         pop_jan = as.integer(pop_jan),
         pop_dec = as.integer(pop_dec)) %>% 
  select(Year, starts_with("pop_"))
```

### Comparison to HMD {.tabset}

BfS vs HMD  

```{r}
compare <- left_join(
  ch_pop_year, 
  hmd_pop_year %>% filter(Country == "Switzerland") %>% rename(HMD_pop = Population)) %>% 
  select(-Country) %>% 
  mutate(diff1 = pop_jan - HMD_pop,
         diff2 = pop_dec - HMD_pop)
```

#### January

Difference between **Jan** pop from BfS and **yearly** pop from HMD

```{r}
summary(compare$diff1)
```

```{r echo=FALSE}
compare %>% 
  ggplot(aes(x = Year, y = diff1)) +
  geom_hline(aes(yintercept = 0), color = "lightblue") +
  geom_vline(aes(xintercept = 1990), color = "lightblue") +
  geom_col() +
  scale_y_continuous(labels = comma, limits = c(-50000, 150000)) +
  theme_bw() +
  xlab("") + ylab("Yearly difference in population estimates")
```

#### Dec 

Difference between **Dec** pop from BfS and **yearly** pop from HMD

```{r}
summary(compare$diff2)
```

```{r echo=FALSE}
compare %>% 
  ggplot(aes(x = Year, y = diff2)) +
  geom_hline(aes(yintercept = 0), color = "lightblue") +
  geom_vline(aes(xintercept = 1990), color = "lightblue") +
  geom_col() +
  scale_y_continuous(labels = comma, limits = c(-50000, 150000)) +
  theme_bw() +
  xlab("") + ylab("Yearly difference in population estimates")
```

```{r include=FALSE}
rm(compare)
```


## Yearly pops with age structure

### Comparison - HMD vs BfS 1  

Data from `px-x-0102030000_101`: *Permanent resident population by sex and age, 1860-2020* [(source)](https://www.bfs.admin.ch/bfs/en/home/statistics/catalogues-databases/data.assetdetail.19105850.html).  With info from docs:

> Reference day: 31 December

```{r}
ch_pop_age_sex <- read.px("data-raw/BfS/px-x-0102030000_101.px") %>% 
  as_tibble() %>% remove_empty() %>% clean_names() %>% 
  filter(alter != "Alter - Total",
         geschlecht == "Geschlecht - Total") %>% 
  rename(Year = jahr) %>% 
  mutate(Year = as.integer(as.character(Year)),
         Age = as.integer(word(alter, 1)),
         Population = as.integer(round(value))) %>% 
  select(-alter, -geschlecht, -value) 
```

We have a bit of discrepancy going on in the data. Example of 2020:  

```{r}
hmd_pop_age_sex %>% 
  filter(Country == "Switzerland") %>% 
  filter(Year == 2020) %>% 
  mutate(Age_cat = cut(Age, c(seq(0, 90, 10),120),
                       include.lowest = TRUE,
                       right = FALSE,
                       labels=c("0-9","10-19","20-29","30-39","40-49",
                                "50-59","60-69","70-79","80-89","90+"))) %>% 
  group_by(Year, Age_cat) %>% 
  summarise(Population = sum(Total)) %>% 
  ungroup() %>% 
  mutate(Source = "HMD") %>% 
  bind_rows(
    ch_pop_age_sex %>% 
      mutate(Age_cat = cut(Age, c(seq(0, 90, 10),120),
                           include.lowest = TRUE,
                           right = FALSE,
                           labels=c("0-9","10-19","20-29","30-39","40-49",
                                    "50-59","60-69","70-79","80-89","90+"))) %>% 
      group_by(Year, Age_cat) %>% 
      summarise(Population = sum(
        # population_on_1_january
        # Total
        # population_on_31_december
        Population
      )) %>% 
      ungroup() %>% 
      mutate(Source = "BfS") 
  ) %>% 
  ggplot(
    aes(x = Age_cat, y = Population / 1000000, 
        fill = factor(Source))) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +
  theme_bw() +
  scale_fill_viridis_d(option = "E", direction = -1) +
  xlab("") + ylab("Population [million]") + 
  labs(title = "Yearly population by age group")
```

### Comparison - HMD vs BfS 2

Using data from `px-x-0102020000_104`: *Demographic balance by age and canton* [(source)](https://www.pxweb.bfs.admin.ch/pxweb/en/px-x-0102020000_104/-/px-x-0102020000_104.px).  

Again 2020 used for examples:  

```{r}
ch_pop_age_sex_2020 <- read_xlsx("data-raw/BfS/px-x-0102020000_104_AGE.xlsx", 
                                 col_types = c("skip", "skip", "skip", 
                                               "skip", "skip", "text", "numeric", 
                                               "numeric"), skip = 2) %>% 
  remove_empty(which = c("rows", "cols")) %>% 
  clean_names() %>% 
  rename(Age = x1) %>% 
  filter(Age != "No indication") %>% 
  mutate(Age = str_replace(Age, fixed(" years"), "")) %>% 
  mutate(Age = str_replace(Age, fixed(" year"), "")) %>% 
  mutate(Age = str_replace(Age, fixed(" or older"), "")) %>% 
  mutate(Age = as.integer(Age)) %>% 
  mutate(Country = "Switzerland", 
         Year = as.integer(2020)) %>% 
  relocate(Country, Year, Age)
```

```{r}
bind_rows(
  hmd_pop_age_sex %>% 
    filter(Country == "Switzerland") %>% 
    filter(Year == 2020) %>% 
    mutate(Age_cat = cut(Age, c(seq(0, 90, 10),120),
                         include.lowest = TRUE,
                         right = FALSE,
                         labels=c("0-9","10-19","20-29","30-39","40-49",
                                  "50-59","60-69","70-79","80-89","90+"))) %>% 
    group_by(Year, Age_cat) %>% 
    summarise(Population = sum(Total)) %>% 
    ungroup() %>% 
    mutate(Source = "HMD"),
  
  ch_pop_age_sex_2020 %>% 
    mutate(Age_cat = cut(Age, c(seq(0, 90, 10),120),
                         include.lowest = TRUE,
                         right = FALSE,
                         labels=c("0-9","10-19","20-29","30-39","40-49",
                                  "50-59","60-69","70-79","80-89","90+"))) %>% 
    group_by(Year, Age_cat) %>% 
    summarise(Population = sum(
      population_on_31_december
    )) %>% 
    ungroup() %>% 
    mutate(Source = "BfS (Dec)"),   
  
  ch_pop_age_sex_2020 %>% 
    mutate(Age_cat = cut(Age, c(seq(0, 90, 10),120),
                         include.lowest = TRUE,
                         right = FALSE,
                         labels=c("0-9","10-19","20-29","30-39","40-49",
                                  "50-59","60-69","70-79","80-89","90+"))) %>% 
    group_by(Year, Age_cat) %>% 
    summarise(Population = sum(
      population_on_1_january
    )) %>% 
    ungroup() %>% 
    mutate(Source = "BfS (Jan)") 
  
) %>% 
  
  ggplot(
    aes(x = Age_cat, y = Population/1000000, 
        fill = factor(Source))) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +
  theme_bw() +
  scale_fill_viridis_d(option = "E", direction = -1) +
  xlab("") + ylab("Population [million]") + 
  labs(title = "Yearly population by age group")
```

<!-- ----------------------------------------------------- -->

# Spain

## Yearly pops 

INE data until 2019. 

```{r}
es_pop_year <- read_xlsx("data-raw/INE/Spain_population.xlsx") %>% 
  remove_empty(which = c("rows", "cols")) %>% 
  clean_names() %>% 
  rename(Year = year) %>% 
  rename(Population = prados_de_la_escosura2020_july) %>% 
  select(Year, Population) %>% 
  mutate(Year = as.integer(Year),
         Population = as.integer(round(Population))) %>% 
  filter(Year != 2020)
```

### Comparison with HMD {.tabset}

Difference between yearly INE and HMD pops.  

```{r}
compare <- left_join(
  es_pop_year %>% rename(INE_pop = Population), 
  hmd_pop_year %>% filter(Country == "Spain") %>% rename(HMD_pop = Population)) %>% 
  select(-Country) %>% 
  mutate(diff = INE_pop - HMD_pop,
         diff_perc = (INE_pop - HMD_pop) / INE_pop)
```

#### Absolute

```{r}
summary(compare$diff)
```

```{r echo=FALSE}
compare %>% 
  ggplot(aes(x = Year, y = diff)) +
  geom_hline(aes(yintercept = 0), color = "lightblue") +
  geom_col() +
  scale_y_continuous(labels = comma, limits = c(-50000, 150000)) +
  theme_bw() +
  xlab("") + ylab("Yearly difference in population estimates")
```

#### Realtive 

```{r}
summary(compare$diff_perc)
```

```{r echo=FALSE}
compare %>% 
  ggplot(aes(x = Year, y = diff_perc)) +
  geom_hline(aes(yintercept = 0), color = "lightblue") +
  geom_col() +
  scale_y_continuous(labels = percent) +
  theme_bw() +
  xlab("") + ylab("% yearly difference in population estimates")
```

## Yearly pops 2020 & 2021

2020 data available from [here](https://www.ine.es/jaxiT3/Tabla.htm?t=31304).^[Needs an update with 2021 if needed for any comparisons!]   

```{r}
es_pop_year <- hmd_pop_year %>% 
  filter(Country == "Spain")

source("R/fn_get_spain_population.R")

es_pop_2020 <- fn_get_spain_population("data-raw/INE/Spain_Population_age-2019.xlsx") %>%
  summarize(Population = sum(Population, na.rm = TRUE))%>%
  mutate(Year = as.integer(2020),
         Population = as.integer(Population)) %>% 
  relocate(Year)

# es_pop_2021 <- ...

es_pop_year %<>% 
  bind_rows(es_pop_2020) %>% 
  # bind_rows(es_pop_2021) %>% 
  fill(Country, .direction = "down")

es_deaths_month %<>% 
  left_join(es_pop_year)
```

```{r include=FALSE}
rm(es_pop_2020, es_pop_2021)
```

```{r echo=FALSE}
es_pop_year %>% 
  ggplot(aes(x = Year, y = Population)) +
  geom_line() +
  theme_bw() +
  xlab("") + ylab("Yearly population")
```

## Pops with age structure 2020 & 2021

```{r}
es_pop_age_2020 <- fn_get_spain_population("data-raw/INE/Spain_Population_age-2019.xlsx") %>%
  mutate(Year = as.integer(2020),
         Age = as.integer(Age),
         Population = as.integer(Population),
         Country = "Spain") %>% 
  rename(Total = Population) %>% 
  relocate(Country, Year)

# es_pop_age_2021 <- ...
```

<!-- ----------------------------------------------------- -->

# Sweden

## 2020 deaths with age & sex structure

Data from SCB.  

```{r}
se_deaths_age_sex_2020 <- read_xlsx("data-raw/SCB/BE0101D9_20210429-200440.xlsx", 
                                    col_types = c("text", "text", "text", "numeric", "numeric"), 
                                    skip = 2) %>% 
  remove_empty(which = c("rows", "cols")) %>% 
  clean_names() %>% 
  filter(!is.na(x2020)) %>% 
  select(-x1, -x2019) %>% 
  mutate(x3 = if_else(x3 == "men", "Male", "Female")) %>% 
  fill(x2) %>% 
  pivot_longer(cols = x2020, names_to = "Year") %>% 
  rename(Age = x2, Sex = x3) %>% 
  mutate(Year = as.integer(str_replace(Year, fixed("x"), "")),
         Age = word(Age, 1),
         Age = str_replace(Age, fixed("100+"), "100"),
         Age = as.integer(Age)) %>% 
  pivot_wider(names_from = Sex, values_from = value) %>% 
  mutate(Male = as.integer(round(Male)),
         Female = as.integer(round(Female))) %>% 
  mutate(Total = Male + Female) %>% 
  arrange(Year, Age) %>% 
  mutate(Country = "Sweden") %>% relocate(Country)
```

### Comparison to HMD results for 2020. {.tabset}

```{r}
compare <- 
  hmd_deaths_age_sex %>% 
  filter(Country == "Sweden" & Year == 2020 ) %>% 
  mutate(Age = ifelse(Age > 100, 100, Age)) %>% 
  select(-Female, -Male) %>% 
  group_by(across(c(-Total))) %>% 
  summarise(Total = sum(Total)) %>% 
  ungroup() %>% 
  rename(Total_hmd = Total) %>% 
  left_join(
    se_deaths_age_sex_2020 %>% 
      mutate(Age = ifelse(Age > 100, 100, Age)) %>% 
      select(-Female, -Male) %>% 
      group_by(across(c(-Total))) %>% 
      summarise(Total = sum(Total)) %>% 
      ungroup() %>% 
      rename(Total_scb = Total) 
  ) %>% 
  mutate(diff = Total_scb - Total_hmd,
         diff_p = ((Total_scb - Total_hmd) / Total_hmd) *100)
```

#### Absolute difference

```{r}
summary(compare$diff)
```

#### Percentage difference

```{r}
summary(compare$diff_p)
```

```{r include=FALSE}
rm(compare)
```

## Yearly pops 

```{r}
se_pop_year <- read_xlsx("data-raw/SCB/be0101_tab9utveng1749-2020.xlsx",
                         skip = 4) %>% 
  remove_empty(which = c("rows", "cols")) %>% 
  clean_names() %>% 
  rename(Population = population) %>% 
  select(year, Population) %>% 
  filter(!is.na(Population)) %>% 
  mutate(year = ifelse(year == "20001)", 2000, year),
         year = as.integer(year),
         Population = as.integer(round(Population))) %>% 
  rename(Year = year) 
```

### Comparison with HMD {.tabset}

SCB vs HMD  

```{r}
compare <- left_join(
  hmd_pop_year %>% filter(Country == "Sweden") %>% rename(HMD_pop = Population),
  se_pop_year %>% rename(SCB_pop = Population)
) %>% 
  select(-Country) %>% 
  mutate(diff = SCB_pop - HMD_pop,
         diff_perc = (SCB_pop - HMD_pop) / SCB_pop)
```

Difference between yearly SCB and HMD pops.  

#### Absolute

```{r}
summary(compare$diff)
```

```{r echo=FALSE}
compare %>% 
  ggplot(aes(x = Year, y = diff)) +
  geom_hline(aes(yintercept = 0), color = "lightblue") +
  geom_col() +
  scale_y_continuous(labels = comma, limits = c(-50000, 150000)) +
  theme_bw() +
  xlab("") + ylab("Yearly difference in population estimates")
```

#### Realtive 

```{r}
summary(compare$diff_perc)
```

```{r echo=FALSE}
compare %>% 
  ggplot(aes(x = Year, y = diff_perc)) +
  geom_hline(aes(yintercept = 0), color = "lightblue") +
  geom_col() +
  scale_y_continuous(labels = percent) +
  theme_bw() +
  xlab("") + ylab("% yearly difference in population estimates")
```

<!-- ----------------------------------------------------- -->

# Computing Environment

`r mu$session()`